---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
#Render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/master/examples>.

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MAGE

<!-- badges: start -->
<!-- badges: end -->

Welcome to MAGE's readme!

MAGE is the Modeller of Allelic Gene Expression, an R package providing extensive functions for various RNAseq-based allelic analyses. This ranges from basic tasks such as (solely) RNAseq-base genotyping, to the analysis of more complex population-level phenomena such as (differential) allelic bias, allellic divergence and (loss of) imprinting analyses. More information about what these are and how to model them using MAGE can be found in the package vignette.

## Installation

MAGE can be installed using the `install_github` function from the `devtools` package:

```{r, warning=FALSE, message=FALSE, eval = FALSE}
library(devtools)
install_github("BioBix/MAGE")
```

## Getting started

MAGE contains a vignette going over its entire anaylis pipeline, which can be found under "articles" on https://biobix.github.io/MAGE.

Both a "regular" and "expanded" vignette are included, the latter writing out all analyses much more in-depth using MAGE's base functions while the former makes use of wrapper functions that handle many analyses and intermediary steps under-the hood. As such, the regular vignette is recommended for first-time users or users that just want a plug-and-play pipeline, while the expanded vignette provides more insight into MAGE's analyses and is recommended when setting out to create your own custom/specialized analysis pipeline.

MAGE has an associated paper pending publication. Coming soon...